A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-12-18 main.tex
Edits to 4 & 5; cover letter; changed title

We begin by comparing the ordinary lasso, PC-lasso, and LMM-lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-lasso, and LMM-lasso was evaluated based on mean squared error of the coefficients (MSE). Suppose $\bbetaHat(\lambda)$ is the vector of coefficients estimated by the lasso with a penalty of $\lambda$, and $\bbeta$ the coefficient values used to simulate the data. Mean squared error is then defined as
\begin{equation}
    \text{MSE} = \mathbb{E}||\bbeta - \bbetaHat(\lambda)||^2_2.
\end{equation}
Figure \ref{fig:mse} presents these results for all three data types in a 1:1 SNR setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental effects), lasso and LMM-lasso perform comparably. PC-lasso results in a slight inflation in MSE compared to the other methods in the simulated data sets due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome.

\begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/beta_mse.png}
    \caption{Estimation accuracy of different methods for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
    \label{fig:mse}
\end{figure}

The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$ (substantial environmental effects). In this case, both PC-lasso and LMM-lasso clearly outperform the ordinary lasso. This improvement is more substantial in the Empirical and Independent Subpopulations data than in the 1D Linear Admixture data. This is due to the fact that the Independent Subpopulations and Empirical data sets are characterized by nonoverlapping genetic subpopulations, meaning that a greater proportion of the environmental effect is in the column space of $\P_{\bX}$. We can quantify the proportion of the environmental component in the column space of $\P_{\bX}$ for each data type using the CP metric defined in \eqref{eqn:ratio} of Section \ref{sec:partition}. Here, we have CP = 0.47, 0.87, and 0.99 for the 1D Linear Admixture, Empirical, and Independent Subpopulations, respectively. The large CP of the Empirical and Independent Subpopulations shows that the environmental effect is more highly correlated with genotype, and therefore has more potential for introducing bias into the estimation of SNP effects; the PC-lasso and LMM-lasso methods correct for this effect, while ordinary lasso does not.  This phenomenon occurs in 1D Linear Admixture setting as well, although the phenomenon is less pronounced because, while the total environmental effect is the same in all three settings, in the admixture setting 53\% of the environmental effect is uncorrelated with genotype and therefore does not introduce bias.

In addition to MSE, these methods' performance was also assessed in terms of variable selection using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Throughout, we use ``true positive'' to mean both that the SNP was selected and that its direction was estimated correctly. Table \ref{tab:var_sel} displays the average TPR for the same settings as in Figure \ref{fig:mse}. The TPRs echo the MSE results in that lasso and LMM-lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the 10 unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-lasso and PC-lasso achieve substantial improvements in TPR over the ordinary lasso. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data.

@@

In order to assess the behavior of these methods in this context, we conducted a simulation based on the Empirical genotype type, but restricted to just two subpopulations, Caucasian and African-American.  1000 SNPs were chosen from the genome-wide Empirical data such that the selected SNPs' absolute difference in MAF between the two subpopulations ranged uniformly from 0 to 0.5. As in the earlier simulations, a 1:1 SNR ($\eta = 0.5$) was used.

\begin{figure}[H]
\centering
\includegraphics[scale = 0.9]{figures/pop_inf_snps_box.png}
\caption{Absolute allele frequency differences of selected, non-causal SNPs across methods, in the presence and absence of environmental confounding for two Empirical subpopulations, dichotomous-discordant effects, $\eta = 0.5$.}
\label{fig:pop_inf}
\end{figure}

Figure \ref{fig:pop_inf} summarizes each method's tendency to select spurious associations driven by differences in MAF between populations.  Even in the absence of environmental effects, the ordinary lasso shows a preference for selecting population-informative SNPs, while PC-lasso and LMM-lasso are reluctant to select such SNPs (recall that with this design, the median MAF difference between populations is 0.25).

In the presence of confounding, however, the lasso shows an overwhelming tendency to select population-informative SNPs.  Its false positive selections are almost always the ones with the largest differences between populations.  PC-lasso and LMM-lasso, on the other, show no such such selection bias.

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/mse_diff_subpops.png}
    \caption{Relative performance of LMM-lasso, PC-lasso for fine and coarse subpopulation structures and varying levels of environmental confounding with Independent Subpopulations data, $\eta = 0.5$, and dichotomous-discordant environmental effects. }
    \label{fig:big_vs_small}
\end{figure}

@@

LMM-lasso-ggmix does tend to provide improved estimates of $\eta$ compared to LMM-lasso-Rakitsch, particularly in the presence of lower $\eta$ values and where $n \ll p$ (Figure \ref{fig:eta_beta_mse}, middle panel), although these improvements did not help it to estimate $\bbeta$ more accurately. In addition, at high signal-to-noise ratios ($\eta=0.8$), the performance of LMM-lasso-ggmix was \anna{highly variable}.

Overall these results show that LMM-lasso-ggmix performs similarly to LMM-lasso-Rakitsch in estimating $\bbeta$ and outperforms LMM-lasso-Rakitsch in estimating $\eta$ when $n \ll p$. However, these are also the settings in which LMM-lasso-ggmix is least computationally efficient. The bottom panel of Figure \ref{fig:eta_beta_mse} shows the computing time required of each method in the various simulation settings, and shows that when $n \ll p$, LMM-lasso-ggmix is up to 60 times slower than LMM-lasso-Rakitsch. In settings where $n \ge p$, LMM-lasso-ggmix is ``faster'' than LMM-lasso-Rakitsch, but this seems to be due to the algorithm getting stuck in a local minimum and prematurely exiting, a phenomenon that also explains the extremely poor estimation accuracy of LMM-lasso-ggmix in the $\eta=0.8$, $n=1400$ setting.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/eta_beta_hat.png}
     \caption{$\bbeta$ and $\eta$ estimation accuracy of LMM-lasso-Rakitsch and LMM-lasso-ggmix for varying $\eta$ levels and sample sizes with Independent Subpopulations data, coarse subpopulation structure, $\xi = 0.8$, and dichotomous-discordant environmental effect structure.}
    \label{fig:eta_beta_mse}
\end{figure}

@@

In this work, we reviewed the concept of population structure in a framework which formally differentiates between effects of population structure and those of the environment. We have provided a broad outline of existing  methods for ameliorating the effects of these unobserved confounding influences. In addition, we have provided a thorough review of two penalized regression approaches that correct for population structure in genetic data: PC-lasso and LMM-lasso.

We have detailed the assumptions made by PC-lasso and LMM-lasso, which model confounding influences as fixed and random effects, respectively. We have also described the ways in which prior analyses of these methods may fail to accurately capture the effects of environmental effects in terms of contributing to selection and estimation bias. We hope that our result partitioning environmental effects into confounding and non-confounded components will be particularly useful in terms of understanding the performance of these approaches.

@@

We considered two implementations of the LMM-lasso method: LMM-lasso-Rakitsch and LMM-lasso-ggmix, which differ in their variance estimation procedures and computational implementations. Although both methods perform similarly in terms of estimating $\bbeta$, the much greater computational efficiency and numerical stability of LMM-lasso-Rakitsch make it a more attractive choice in most scenarios, although LMM-lasso-ggmix may be useful if the objective of the analysis is to accurately estimate narrow-sense heritability.

>>>>

We begin by comparing the ordinary lasso, PC-lasso, and LMM-lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-lasso, and LMM-lasso was evaluated based on mean squared error (MSE) of the coefficients, MSE = $\mathbb{E}||\bbeta - \bbetaHat(\lambda)||^2_2$, where $\bbetaHat(\lambda)$ is the vector of coefficients estimated by the lasso with a penalty of $\lambda$, and $\bbeta$ the true coefficient values.
Figure \ref{fig:mse} presents these results for all three data types in a 1:1 SNR setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental effects), lasso and LMM-lasso perform comparably. PC-lasso results in a slight inflation in MSE compared to the other methods in the simulated data sets due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome.

\begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/beta_mse.png}
    \caption{Estimation accuracy of different methods for different data types in the presence and absence of environmental confounding with 4 subpopulations.}
    \label{fig:mse}
\end{figure}

The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$ (substantial environmental effects). In this case, both PC-lasso and LMM-lasso clearly outperform the ordinary lasso. This improvement is more substantial in the Empirical and Independent Subpopulations data than in the 1D Linear Admixture data. This is because there is less overlap between subpopulations than in the admixture data, meaning that a greater proportion of the environmental effect is in the column space of $\P_{\bX}$. We can quantify this proportion using the CP metric defined in \eqref{eqn:ratio}. Here, we have CP = 0.47, 0.87, and 0.99 for the 1D Linear Admixture, Empirical, and Independent Subpopulations, respectively. The large CP of the Empirical and Independent Subpopulations shows that the environmental effect is more highly correlated with genotype, and therefore has more potential for introducing bias into the estimation of SNP effects; the PC-lasso and LMM-lasso methods correct for this effect, while ordinary lasso does not.  This phenomenon occurs in 1D Linear Admixture setting as well, although the phenomenon is less pronounced because, while the total environmental effect is the same in all three settings, in the admixture setting 53\% of the environmental effect is uncorrelated with genotype and therefore does not introduce bias.

In addition to MSE, these methods' performance was also assessed in terms of variable selection using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Throughout, we use ``true positive'' to mean both that the SNP was selected and that its direction was estimated correctly. Table \ref{tab:var_sel} displays the average TPR for the same settings as in Figure \ref{fig:mse}. The TPRs echo the MSE results in that lasso and LMM-lasso perform comparably in the absence of environmental confounding across all data types, with PC-lasso slightly less accurate at identifying causal SNPs. In the presence of environmental confounding, both LMM-lasso and PC-lasso achieve substantial improvements in TPR over the ordinary lasso. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data.

@@

In order to assess the behavior of these methods in this context, we conducted a simulation using Empirical genotypes, but restricted to just two subpopulations, Caucasian and African-American.  A total of 1000 SNPs were chosen such that the selected SNPs' absolute difference in MAF between the two subpopulations ranged uniformly from 0 to 0.5. As in the earlier simulations, a 1:1 SNR ($\eta = 0.5$) was used to simulate the phenotype.

\begin{figure}[hbt]
\centering
\includegraphics[scale = 0.9]{figures/pop_inf_snps_box.png}
\caption{Absolute allele frequency differences of selected, non-causal SNPs across methods, in the presence and absence of environmental confounding for two Empirical subpopulations, $\eta = 0.5$.}
\label{fig:pop_inf}
\end{figure}

Figure \ref{fig:pop_inf} summarizes each method's tendency to select spurious associations driven by differences in MAF between populations.  Even in the absence of environmental effects, the ordinary lasso shows a preference for selecting population-informative SNPs, while PC-lasso and LMM-lasso are reluctant to select such SNPs (recall that with this design, the median MAF difference between populations is 0.25).  In the presence of confounding, however, the lasso shows an overwhelming tendency to select population-informative SNPs.  Its false positive selections are almost always the ones with the largest differences between populations.  PC-lasso and LMM-lasso, on the other, show no such such selection bias.

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/mse_diff_subpops.png}
    \caption{Relative performance of LMM-lasso, PC-lasso for fine and coarse subpopulation structures and varying levels of environmental confounding with Independent Subpopulations, $\eta = 0.5$. }
    \label{fig:big_vs_small}
\end{figure}

@@

LMM-lasso-ggmix does tend to provide improved estimates of $\eta$ compared to LMM-lasso-Rakitsch, particularly in the presence of lower $\eta$ values and where $n \ll p$ (Figure \ref{fig:eta_beta_mse}, middle panel), although these improvements did not help it to estimate $\bbeta$ more accurately. In addition, at high signal-to-noise ratios ($\eta=0.8$), the performance of LMM-lasso-ggmix was highly variable.

Overall these results show that LMM-lasso-ggmix performs similarly to LMM-lasso-Rakitsch in estimating $\bbeta$ and outperforms LMM-lasso-Rakitsch in estimating $\eta$ when $n \ll p$. However, these are also the settings in which LMM-lasso-ggmix is least computationally efficient. The bottom panel of Figure \ref{fig:eta_beta_mse} shows the computing time required of each method in the various simulation settings, and shows that when $n \ll p$, LMM-lasso-ggmix is up to 60 times slower than LMM-lasso-Rakitsch. In settings where $n \ge p$, LMM-lasso-ggmix is ``faster'' than LMM-lasso-Rakitsch, but this seems to be due to the algorithm getting stuck in a local minimum and prematurely terminating, a phenomenon that also explains the extremely poor estimation accuracy of LMM-lasso-ggmix in the $\eta=0.8$, $n=1400$ setting.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/eta_beta_hat.png}
     \caption{Estimation accurate of LMM-lasso-Rakitsch and LMM-lasso-ggmix for $\bbeta$ and $\eta$ for varying $\eta$ levels and sample sizes with Independent Subpopulations data, coarse subpopulation structure, $\xi = 0.8$.}
    \label{fig:eta_beta_mse}
\end{figure}

@@

In this work, we reviewed the concept of population structure in a framework which formally differentiates between effects of population structure and those of the environment. We have provided a broad outline of existing  methods for ameliorating the effects of these unobserved confounding influences. In addition, we have provided a thorough review of two penalized regression approaches that correct for population structure in genetic data: PC-lasso and LMM-lasso. We have detailed the assumptions made by PC-lasso and LMM-lasso, which model confounding influences as fixed and random effects, respectively. We have also described the ways in which prior analyses of these methods may fail to accurately capture the effects of environmental effects in terms of contributing to selection and estimation bias. We hope that our result partitioning environmental effects into confounding and non-confounded components will be particularly useful in terms of understanding the performance of these approaches.

@@

We considered two implementations of the LMM-lasso method: LMM-lasso-Rakitsch and LMM-lasso-ggmix, which differ in their variance estimation procedures and computational implementations. Although both methods perform similarly in terms of estimating $\bbeta$, the much greater computational efficiency and numerical stability of LMM-lasso-Rakitsch make it a more attractive choice in most scenarios, although LMM-lasso-ggmix may be useful if the objective of the analysis is to accurately estimate variance components.
